---
title: "Project"
author: "Fonti Kar & Emi Tanaka"
date: 2025/07/14
institute: Australian National University
execute:
  cache: true
  echo: true
format: 
  anu-light-revealjs:
    width: 1920
    height: 1080
    disable-layout: false
    auto-stretch: false
    html-math-method: katex
    code-block-height: 1000px
    css: [/assets/tachyons-addon.css]
---

## AAGI Project Structure

```{r}
#| echo: false
options(width = 70)
```


- Multi-environment trials (METs) are field experiments conducted in multiple locations and/or seasons to evaluate the performance of different crop genotypes. 
- Trial location and season/year is used synonymously with "environment"

::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| fig-height: 7
#| cache: false
library(cropscan)
plot_field(met2_pheno)
```

:::

::: {.column width="50%"}

- $G = 191$ genotypes tested across all $E = 5$ environments with $n = 288$ observations and $B = 2$ blocks per trial.

```{r}
#| echo: true
concurrence_matrix(met2_pheno)
```


:::
:::


## Key takeaways {background-color="#F5EDDE"}

- Project structure gives you and your team an unspoken understanding
  - Keeps your kitchen organised
- This gives makes your work intuitive and efficient
- A solid project structure sets the foundation for more reproducible workflows





